Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2024-01-25 manu-draft.qmd
edits from KFW

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable to, if not more significant than, other interconnected critical coastal environments such as mangroves and salt marshes [@duarte2008a <!--# any more recent references? -->], which threatens the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; <!--# Add Hall et al. 1999, Florida bay seagrass dieoff and recovery --> @orth2006; @burkholder2007]. Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Contemporary management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

@@

Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and the second largest in the Gulf of Mexico. The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers that contribute to a majority of freshwater inflow to the bay. Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer<!--# reference? -->. The watershed is heavily developed and includes over 3 million people <!--# reference economic valuation? -->with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023]. The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022a]. Important subtidal habitats include seagrasses, hard bottom, tidal flats, and oyster reefs, where a majority of management efforts have historically focused on restoring and maintaining seagrass cover [@sherwood2017]. Additional native habitats include intertidal wetlands (mangroves, salt marshes, salt barrens) and pine forests, oak hammocks, and freshwater wetlands present in upland habitats [@Robison20]. Losses of native uplands and potentially restorable habitats to development in the watershed from 1990 to 2020 have been estimated at 188,429 acres [76,254 ha, @beck2023].

@@

Additional limitations of our model may relate to an incomplete description of factors influencing seagrass growth. This relates to the inclusion of additional drivers as well as issues related to correlation/causation. For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth [@greening2014; @sherwood2017]. As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity. Our use of time period (pre-/post-2016) as a categorical variable partially addressed this issue. Modeling seagrass change as related to temperature or salinity for the entire record would have shown a spurious correlation of both with seagrass given the long-term recovery of seagrass. Thus, time period was necessary to control for these confounding relationships. Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay [OTB in particular where the most losses have been seen, @tbep0696; @Beck20a]. This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity. A final consideration for our models relates to how seagrasses may influence their environment, particularly for the EPC and FIM <!--# Should be FIM here, correct? --> datasets where temperature and salinity were measured at the same locations as seagrass. For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses. This explanation cannot be ruled out with the existing datasets, although the trend analyses and models suggest that climate-related stressors driving seagrass loss is a more likely scenario.

@@

Additional biotic factors could be influencing regional patterns in seagrass growth. In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern \[ <!--# add Indian River paper? -->@jei2022; @scolaro2023\]. Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay [@Beck20b]. Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997], in addition to insufficient data in Tampa Bay to clearly describe growth patterns and nutrient cycling related to macroalgae. Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth [@beck2022a; @scolaro2023; @tomasko2023], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years. Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018]. For example, the parasitic slime mold *Labryinthula* spp. that causes seagrass wasting disease has been known to infect *Thalassia testudinium* in Tampa Bay [@blakesley2001], although it is unclear if these infections have had large-scale, population-level effects. Existing research has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].

>>>>

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable to, if not more significant than, other interconnected critical coastal environments such as mangroves and salt marshes [@duarte2008a], which threatens the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @hall1999; @orth2006; @burkholder2007]. Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Contemporary management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

@@

Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and the second largest in the Gulf of Mexico. The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers that contribute to a majority of freshwater inflow to the bay. Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer [@morrison2006]. El Ni\\\~{n}o weather events have also been associated with prolonged periods of heavy rainfall contributing to seagrass reductions through increased stormwater nutrient loads [@greening2006; @morrison2006]. The watershed is heavily developed and includes over 3 million people <!--# reference economic valuation? -->with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023]. The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022a]. Important subtidal habitats include seagrasses, hard bottom, tidal flats, and oyster reefs, where a majority of management efforts have historically focused on restoring and maintaining seagrass cover [@sherwood2017]. Additional native habitats include intertidal wetlands (mangroves, salt marshes, salt barrens) and pine forests, oak hammocks, and freshwater wetlands present in upland habitats [@Robison20]. Losses of native uplands and potentially restorable habitats to development in the watershed from 1990 to 2020 have been estimated at 188,429 acres [76,254 ha, @beck2023].

@@

Additional limitations of our model may relate to an incomplete description of factors influencing seagrass growth. This relates to the inclusion of additional drivers as well as issues related to correlation/causation. For the former, the primary management paradigm in Tampa Bay for the past three decades has relied on the role of nutrients, nitrogen in particular, in affecting light environments for seagrass growth [@greening2014; @sherwood2017]. As such, light attenuation or water clarity could have been included in our models to more completely describe factors influencing growth, i.e., the residual differences after accounting for light attenuation could additionally be explained by temperature or salinity. Our use of time period (pre-/post-2016) as a categorical variable partially addressed this issue. Modeling seagrass change as related to temperature or salinity for the entire record would have shown a spurious correlation of both with seagrass given the long-term recovery of seagrass. Thus, time period was necessary to control for these confounding relationships. Additionally, light attenuation has been relatively consistent since 2016 and within the limits estimated to be supportive of seagrass growth in Tampa Bay [OTB in particular where the most losses have been seen, @tbep0696; @Beck20a]. This suggests that other factors are influencing seagrass growth and the inclusion of light attenuation in the models was likely not critical for describing the roles of temperature and salinity. A final consideration for our models relates to how seagrasses may influence their environment, particularly for the PDEM and FIM datasets where temperature and salinity were measured at the same locations as seagrass. For example, temperature may simply be higher in locations where seagrasses are present, i.e., seagrasses may be influencing their environment rather than the environment influencing seagrasses. This explanation cannot be ruled out with the existing datasets, although the trend analyses and models suggest that climate-related stressors driving seagrass loss is a more likely scenario.

@@

Additional biotic factors could be influencing regional patterns in seagrass growth. In Tampa Bay and elsewhere, the presence of macroalgal species and how these may influence seagrass growth and recovery has been a concern [@hall2022; @jei2022; @scolaro2023]. Macroalgae abundance has increased over time and has been observed to colonize locations where seagrass was formerly present in Tampa Bay [@Beck20b]. Competitive differences between seagrasses and macroalgae are poorly understood in these systems [but see @bell1997], in addition to insufficient data in Tampa Bay to clearly describe growth patterns and nutrient cycling related to macroalgae. Discrete pollutant loading events in Tampa Bay have been documented to promote macroalgae growth [@beck2022a; @scolaro2023; @tomasko2023], whereas the role of chronic nutrient loading and changing climatic conditions in promoting growth and interactions with seagrasses is less understood, particularly in recent years. Finally, additional research has focused on how diseases and pathogens can influence seagrass growth patterns in Florida [@robblee1991; @duffin2021; @vanbogaert2018]. For example, the parasitic slime mold *Labryinthula* spp. that causes seagrass wasting disease has been known to infect *Thalassia testudinium* in Tampa Bay [@blakesley2001], although it is unclear if these infections have had large-scale, population-level effects. Existing research has primarily focused on describing spatial patterns, past die-off events, or immunology of these pathogens [@robblee1991; @duffin2021].
